---
title: "Project: Global Fish Watch Vessel Data"
format: html
execute: 
  echo: false
  warning: false
  message: false
embed-resources: true
---

### Data Description

The data used was acquired through Global Fish Watch.You can view the original data source [here](https://globalfishingwatch.org/data-download/datasets/public-training-data-v1). Global Fishing Watch. 2020. Anonymized AIS training data, Version 1.0.

### Data Cleaning

In order to clean the data I had to first combing all my data sets into one. I created a new column that combined the gear type. I then cleared out all the data that was invalid. Finally I summarized and switched my data

### Visualization 1: Average Vesses Speed by Gear Type

This chart shows how the average speed of fishing vessels differs depending on the gear they use. It uses a boxplot to show the range and average speeds for each type of vessel.

::: center
![Boxplot comparing vessel speed across four gear types. The chart shows purse seines as the fastest on average, followed by pole and line, trawlers, and trollers.](average_vessel_speed.pdf)
:::

### Visualization 2: Time Spent Fishing by Gear Type

This bar chart shows how much time, on average, vessels using each gear type spend fishing. It compares the proportion of time spent fishing to total time.

::: center
![Bar chart showing average percent of time spent fishing. Trollers spend the most time fishing, followed by trawlers, purse seines, and then pole and line.](average_time_fishing_by_gear.pdf)
:::

```{r}
library(tidyverse)
library(dplyr)
library(lubridate)
```

```{r}
gear_files <- list.files("fishing_data", full.names = TRUE)

gear_labels <- basename(gear_files) %>% str_remove(".csv")

fishing_data <- map2_dfr(
  gear_files,       # list of CSV file paths
  gear_labels,      # corresponding gear type labels
  ~ read_csv(.x, show_col_types = FALSE) %>%  # read each file
      mutate(gear_type = .y)                  # tag with gear type
)

```

```{r}
fishing_data_clean <- fishing_data %>%
  # Remove rows where is_fishing is -1 (invalid/missing)
  filter(is_fishing != -1) %>%
  
  # Convert Unix timestamp to POSIXct datetime format
  mutate(
    timestamp = as.POSIXct(timestamp, origin = "1970-01-01", tz = "UTC"),
    
    # Convert is_fishing scores to binary (0 = not fishing, 1 = fishing)
    is_fishing = ifelse(is_fishing > 0, 1, 0)
  ) %>%
  
  # Keep only relevant columns for analysis
  select(mmsi, timestamp, speed, distance_from_shore, distance_from_port, is_fishing, gear_type)
```

```{r}
# Group by vessel and gear type and compute summary statistics
summary_by_vessel <- fishing_data_clean %>%
  group_by(mmsi, gear_type) %>%  # Group by vessel and gear type
  summarize(
    avg_speed = mean(speed, na.rm = TRUE),                 # Average vessel speed
    avg_dist_shore = mean(distance_from_shore, na.rm = TRUE),  # Average distance from shore
    avg_dist_port = mean(distance_from_port, na.rm = TRUE),    # Average distance from port
    pct_fishing = mean(is_fishing, na.rm = TRUE) * 100,    # Percent of time fishing
    .groups = "drop"                                       # Ungroup after summarize
  )

```

```{r}
# Create a summary table of average fishing time by gear type
summary_by_gear <- summary_by_vessel %>%
  group_by(gear_type) %>%                       # Group by gear type
  summarize(
    mean_pct_fishing = mean(pct_fishing, na.rm = TRUE),  # Average % time fishing
    .groups = "drop"
  )
```

```{r}
avgspd_gear <- ggplot(summary_by_vessel, aes(x = gear_type, y = avg_speed)) +
  geom_boxplot() +  # Draw a boxplot
  labs(
    title = "Average Vessel Speed by Gear Type",
    x = "Gear Type",
    y = "Speed (knots)"
  ) +
  theme_minimal()   # Use a clean theme


ggsave("average_vessel_speed.pdf", plot = avgspd_gear, width = 7, height = 5, dpi = 300)
```

```{r}

TimeSpent_Fish <- ggplot(summary_by_gear, aes(x = reorder(gear_type, -mean_pct_fishing), y = mean_pct_fishing)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Average Time Spent Fishing by Gear Type",
    x = "Gear Type",
    y = "% Time Fishing"
  ) +
  theme_minimal()

# Then, save the plot as a PNG file
ggsave("average_time_fishing_by_gear.pdf", plot = TimeSpent_Fish, width = 7, height = 5, dpi = 300)
```
